---
title: "Working Script Practical 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r}
library(mHMMbayes)
source('./helpers.R')
```

## Load Data

```{r}
emotion_data <- readRDS("./data/data_Rowland2020.rds")
emotion_data <- emotion_data[,c('subj_id', 'dayno', 'beep','happy','relaxed','anxious','depressed')] # only use a subset of variables
```

```{r}
emotion_data$subj_id2[1] <- 1
for(j in 2:nrow(emotion_data)){
  if(emotion_data$subj_id[j] == emotion_data$subj_id[j-1]){
    emotion_data$subj_id2[j] <- emotion_data$subj_id2[j-1]
  } else {
    emotion_data$subj_id2[j] <- emotion_data$subj_id2[j-1] + 1
  }
}
emotion_data$subj_id <- emotion_data$subj_id2
emotion_data <- subset(emotion_data, select = -c(subj_id2))
```

**Exercise 1**

**A.** Inspect the first few rows of the emotion data.
```{r}

```

**B.** Inspect the data over time using visualization. Most useful is visualizing time sequences for subjects separately, e.g., by using the option `facet_wrap(vars(subj_id))` in ggplot, or density plots.

```{r, warning = FALSE, message = FALSE}
library(ggplot2)

emotion_data$beep2 <- (emotion_data$dayno - 1) * 6 + emotion_data$beep # add a beep number

# Create a long format data frame for ggplot
ggplot_emotion <- rbind(data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$happy,
                                 emotion = "happy"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$relaxed,
                                 emotion = "relaxed"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$anxious,
                                 emotion = "anxious"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$depressed,
                                 emotion = "depressed"))

ggplot_emotion$emotion <- factor(ggplot_emotion$emotion, levels = c("happy", "relaxed", "anxious", "depressed")) # turn the emotion variable into a factor

ggplot(ggplot_emotion[ggplot_emotion$ID %in% c(1:5),]) + # only plot the first 5 subjects
  geom_line(mapping = aes(x = beep2, y = outcome, color = emotion)) + # line plot for time series (different colors for each emotion)
  facet_wrap(vars(ID), ncol = 1) + # plot separate panels for different subjects
  scale_color_manual(values = c("happy" = "darkgreen", "relaxed" = "lightgreen", "anxious" = "darkred", "depressed" = "red")) # color palette

ggplot(ggplot_emotion, aes(x = outcome)) +
  geom_density() +
  facet_wrap(~emotion) # plot separate panels for different emotions
```

## Missing Data and Night Gap

**Exercise 2**
Use the `insert_nightgap()` helper function to insert 8 rows of missing values for each subject for each night gap. Call the new dataset `emotion_nightgap`.

```{r}
emotion_nightgap <- insert_nightgap(data = emotion_data, # make sure the first column contains the subject IDs, and the second column contains the day numbers
                                    beeps_per_night = 8, # we want to insert 8 beeps for each night
                                    col_id = 1, # the first column contains the subject IDs
                                    col_day = 2 # the second column contains the day numbers
                                    )
```

## Setting Up the Multilevel Hidden Markov Model

**Exercise 3**
Create a dataset `emotion_mHMM` which has the subject ID variable in the first column, followed by only the affective variables you want to use in the hidden Markov model. 

```{r}

```

### Starting Values

**Exercise 4**

**A.** Explore the data to determine starting values for the emission distributions.

```{r}
# Determine observed range
apply(emotion_mHMM[, 2:5], 2, range, na.rm = TRUE) # observed range is from 0 to 100 for all variables

apply(emotion_mHMM[, 2:5], 2, function(x){
  x_low <- x[x <= 50] # lower range
  x_high <- x[x > 50] # larger range
  c(median(x_low, na.rm = TRUE), median(x_high, na.rm = TRUE))
})

## The lower proposed starting values for anxious and depressed are quite close to the boundary of 0, so we should bring the starting values for state 1 and 2 closer together for these variables.
```

**B** Now, use the obtained values to create input for the model to use as starting values. First, set up the model input arguments: `m` for the number of states the model should infer, `n_dep` for the number of dependent variables used for fitting the model, and then the starting values for the emission distribution, saved in the object `start_emiss`. `start_emiss` should be a list with `n_dep` elements. Each element within the list should be a matrix with 2 columns and `m` rows. The means are contained in the first column of this matrix, and the standard deviations are contained in the second column of the matrix.

```{r}
# setting up input parameters
m <- 2
n_dep <- 4

start_emiss <- list(matrix(c(..., ...,                                     # happy
                             ..., ...), byrow = TRUE, ncol = 2, nrow = m),
                    ...,                                     # relaxed,
                    ...,                                     # anxious 
                    ...                                     # depressed)
```

**C.** Set up the starting values for the transition probabilities in the object `start_gamma` using the starter code below.

```{r}
start_gamma <- matrix(c(..., ...,
                        ..., ...), byrow = TRUE, ncol = m, nrow = m)
```

### Specifying Priors

**Exercise 5**
Use the `help()` function to inspect the function `prior_emiss_cont()` and its arguments.

```{r}

```

**Exercise 6**
Set up a weakly informative prior for the continuous emission distribution using the function `prior_emiss_cont()` and save the object in `emotion_prior_emiss`. Keep the above considerations in mind. You will only have to fill in the dots below.

```{r}
# specifying weakly informative prior for continuous emission distributions

emotion_prior_emiss <- prior_emiss_cont(
  gen = list(m = m, n_dep = n_dep),
  emiss_mu0 = list(...,  # happy
                   ...,  # relaxed
                   ...,  # anxious
                  ...),  # depressed
  emiss_K0 = ...,
  emiss_V = ...,
  emiss_nu = ...,
  emiss_a0 = ...,
  emiss_b0 = ...,
)
```

## Fitting the Multilevel Hidden Markov Model

**Exercise 7**
Fit a 2 state multilevel hidden Markov model using the function `mHMM()`, and save your fitted model in an object named `out_2st_emotion`. See below for the general structure of the `mHMM()` function, you only have to fill in the dots. The provided code assumes continuous observations, no level 2 covariates and only a prior on the Normal emission distribution.

```{r}
# Run a model without covariate(s) and default priors:
set.seed(...) ## choose a seed for reproducibility
out_2st_emotion <- mHMM(s_data = ...,
                data_distr = "continuous",
                gen = list(m = ..., n_dep = ...),
                start_val = ...,
                mcmc = list(J = 500, burn_in = 200))
```

## Inspecting general model output 

**Exercise 8**
Inspect the global output by using the inbuilt `print()` and `summary()` options for a class `mHMM` object created by using the function `mHMM()` within the `mHMMbayes` package. How many subjects were included in the analysis? On average, do students remain within one and the same state for a long time? What mood do each of the states represent?

```{r}

```

**Exercise 9**
Save the group level transition probabilities gamma in an object named `gamma_group`, and the group level emission distribution parameters in an object named `emiss_group`, using the functions `obtain_gamma()` and `obtain_emiss()`, respectively. Inspect the objects.

```{r}

```

## Visualizing the Obtained Output

Load the *ggmHMM* package
```{r}
devtools::install_github("pepijnvink/ggmHMM") # for visualizations using ggplot2
library(ggmHMM)
```

**Exercise 10**
Visualize the group level transition probabilities gamma using a heatmap. You can use the `ggmHMM` function `plot_gamma()`.

```{r}
plot_gamma(model = gamma_group) # provide either a model object made with `mHMM()` or a transition matrix obtained with `obtain_gamma()`
```

**Exercise 11**
Visualize the group level emission distributions. Various visualizations are possible, for example bar plots depicting the emission distribution means over states and dependent variables (ussing the *ggmHMM* function `plot_emiss()`), or depicting the densities of the normal emission distributions (using the helper function `plot_density()`), combining information on both the mean and standard deviation.

```{r}
# Bar chart using the ggmHMM package
plot_emiss(model = out_2st_emotion, subject_effects = FALSE) # bar chart

# density plot using the helper function
plot_dens(model = out_2st_emotion) # density plot
```

## Obtaining the most likely sequence of hidden states

**Exercise 12**
Using the function `vit_mHMM()`, the object `out_2st_emotion` containing our fitted multilevel HMM model, and the data `emotion_mHMM`, obtain the most likely hidden state sequence. Save the state sequences in the object `emotion_states_2st`, and inspect the object.

```{r}

```

**Exercise 13**
Use the *ggmHMM* function `plot_viterbi()` to plot the inferred state sequence over a time for a selection of the subjects within the data.

```{r}

```

